---
title: "Supplemental Materials"
subtitle: "*How does adversity relate to performance across different abilities in the same person?*"
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    fig.path: "figures/"
format: 
  docx:
    reference-doc: reference-doc.docx
    fig-dpi: 300
    fig-format: pdf
  gfm: 
    fig-format: jpeg
    fig-dpi: 300
bibliography: ../../bib-files/references.bib
csl: ../../bib-files/apa.csl
---

```{r}
#| label: setup
#| include: false

# Load libraries
library(tidyverse)
library(patchwork)
library(flextable)

# Load staged results
load("r-objects.Rdata")

# load custom functions
source("../../scripts/0-corr_table.R")

# load data for reviewer request
load("../../data/seccyd-dvs-wj.Rdata")

# ggplot2 theme
theme_set(
  theme_bw() +
    theme(
      axis.line.y       = element_line(),
      axis.text.y       = element_text(size = rel(1.1)),
      axis.title.y      = element_text(size = rel(1.25), margin = margin(1,0,0,0,"lines")),
      axis.ticks.y      = element_line(),
      axis.text.x       = element_text(size = rel(1.1)),
      axis.title.x      = element_text(size = rel(1.25), margin = margin(1,0,0,0,"lines")),
      axis.line.x       = element_line(),
      panel.border      = element_blank(), 
      panel.spacing.y   = unit(0.5, "lines"),
      plot.margin       = margin(.25,.25,.25,.25,"lines"),
      plot.background   = element_rect(color = NA),
      plot.title        = element_text(size = rel(1.25), hjust = 0.5, margin = margin(0,0,.5,0, "lines")),
      plot.subtitle     = element_blank(),
      panel.grid        = element_line(color = NA),
      strip.background  = element_blank(), 
      strip.placement   = "outside",
      strip.text        = element_text(size = rel(1), angle = 0)
    )
)

## Table settings
# set up flextable for tables
set_flextable_defaults(
  font.family = "Times", 
  font.size = 10,
  font.color = "black",
  line_spacing = 1,
  padding.bottom = 1, 
  padding.top = 1,
  padding.left = 1,
  padding.right = 1
)
```

This document contains three parts. First, we provide correlations and descriptive statistics for all indicators used to create aggregate variables for harshness and unpredictability. Second, we provide full regression tables for both primary and secondary analyses. Third, we provide figures for both primary and secondary analyses where WJ scores are uncentered. Note that these analyses are the same, however, centering WJ scores allows for better visual comparison of each subtest slope. The uncentered scores retain visual information about mean differences in performance across subtest types.

Below are links to each of these sections:

- [Correlations Among All Harshness and Unpredictability Indicators](#stab1)
- **Full Primary Analysis Results**
  - [Family Income Disadvantage](#p1)
  - [Neighborhood Socioeconomic Disadvantage](#p2)
  - [Family Transitions Results](#p3)
  - [Neighborhood Soocioeconomic Variability](#p4)
- **Full Secondary Analysis Results**
  - [Family Income Standard Deviation](#s1)
  - [Family Income Residual Standard Deviation](#s2)
  - [Family Income Average Percent Change](#s3)
- **Figures with Uncentered WJ Scores**
  - [Primary Analyses](#sfig1)
  - [Secondary Analyses](#sfig2)
- **Reviewer Requested**
  - [WJ Subtest Correlations Across Time](#srev1)

{{< pagebreak >}}

## Correlations Among All Harshness and Unpredictability Indicators {#stab1}

```{r}
#| tab.id: stable1
#| results: markup

stable_ivs1 |> 
  flextable() |> 
  fontsize(size = 8) |> 
  autofit() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  add_header_row(
    values = " ",
    colwidths = 9
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 1a. Correlations and descriptive statistics among all harshness indicators.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 9
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 9
  ) 
```

```{r}
#| tab.id: stable1
#| results: markup

stable_ivs2 |> 
  flextable() |> 
  fontsize(size = 8) |> 
  autofit() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  add_header_row(
    values = " ",
    colwidths = 15
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 1b. Correlations and descriptive statistics among all unpredictability indicators.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 15
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 15
  ) 
```

{{< pagebreak >}}

## Full Family Income Disadvantage Results {#p1}

```{r}
#| tab.id: stable2
#| results: markup

stables_primary[[2]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 2. Full model results for Family Income Disadvantage.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable3
#| #| results: markup

stables_primary[[2]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 3. Simple Effects for Family Income Disadvantage.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Full Neighborhood Disadvantage Results {#p2}

```{r}
#| tab.id: stable4
#| results: markup

stables_primary[[4]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 4. Full model results for Neighborhood Socioeconomic Disadvantage.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable5
#| results: markup

stables_primary[[4]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 5. Simple Effects for Neighborhood Socioeconomic Disadvantage.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Full Family Transitions Results {#p3}

```{r}
#| tab.id: stable6
#| results: markup

stables_primary[[1]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 6. Full model results for Family Transitions.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable7
#| results: markup

stables_primary[[1]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 7. Simple Effects for Family Transitions.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Full Neighborhood Socioeconomic Variability

```{r}
#| tab.id: stable8
#| results: markup

stables_primary[[5]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 8. Full model results for Neighborhood Socioeconomic Variability.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable9
#| results: markup

stables_primary[[5]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 9. Simple Effects for Neighborhood Socioeconomic Variability.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Full Family Income Variability (SD) {#s1}

```{r}
#| tab.id: stable10
#| results: markup

stables_secondary[[2]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 10. Full model results for Family Income Variability (SD).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable11
#| results: markup

stables_secondary[[2]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 11. Simple Effects for Family Income Variability (SD).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Full Family Income Variability (Residual SD) {#s2}

```{r}
#| tab.id: stable12
#| results: markup

stables_secondary[[3]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 12. Full model results for Family Income Variability (Residual SD).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable13
#| results: markup

stables_secondary[[3]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 13. Simple Effects for Family Income Variability (Residual SD).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}


## Full Family Income Variability (Average Percent Change) {#s3}

```{r}
#| tab.id: stable14
#| results: markup

stables_secondary[[1]] |> 
  select(parameter, b, B, p, rope_p) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |> 
  compose(i = 1, j = c(2), as_paragraph(as_i("b ")), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("B "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(4), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(5), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 14. Full model results for Family Income Variability (Average Percent Change).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  autofit()
```

```{r}
#| tab.id: stable15
#| results: markup

stables_secondary[[1]] |> 
  select(parameter, slope, slope_p, rope_slope_p) |> 
  filter(row_number() > 4) |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  compose(i = 1, j = c(1), as_paragraph(""), part = "header") |>  
  compose(i = 1, j = c(2), as_paragraph(as_i("b "), "[95% CI]"), part = "header") |>  
  compose(i = 1, j = c(3), as_paragraph(as_i("p")), part = "header") |>
  compose(i = 1, j = c(4), as_paragraph("ROPE ", as_i("p")), part = "header") |>
  add_header_row(
    values = " ",
    colwidths = 4
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Supplemental Table 15. Simple Effects for Family Income Variability (Average Percent Change).")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 4
  ) |> 
  autofit()
```

{{< pagebreak >}}

## Supplemental Figure 1 {#sfig1}

```{r}
#| label: sfigure1
#| fig-width: 6
#| fig-height: 7.5

sfig1
```

## Supplemental Figure 2 {#sfig2}

```{r}
#| label: sfigure2
#| fig-width: 5.5
#| fig-height: 7.5

sfig2
```

{{< pagebreak >}}

## WJ Subtest Correlations Across Time (Reviewer Requested) {#srev1}

```{r}
wj_order <- 
  c(
    "wj_pscmp",
    "wj_picvo",
    "wj_calc",
    "wj_vrba",
    "wj_lwid",
    "wj_memse",
    "wj_appld",
    "wj_incom",
    "wj_wrdat",
    "wj_memna"
  )

wj_labels <- 
  c(
    "Passage Comprehension",
    "Picture Vocab",
    "Calculations",
    "Verbal Analogies",
    "Letter-Word Pronunciation",
    "Short-Term Memory",
    "Applied Problems",
    "Auditory Processing",
    "Unfamiliar Words",
    "Auditory-Visual Associations"
  )

assess_fct <- 
  c(
    "54m", 
    "1g", 
    "3g", 
    "5g", 
    "15y"
  )

assess_label <- 
  c(
    "54 month",
    "Grade 1",
    "Grade 3",
    "Grade 5",
    "Age 15"
  )

wj_corrs_time <- 
  map2(
    wj_order, wj_labels,
    function(x, y){
      seccyd_dvs_wj_data2 |> 
        select(id, assessment, any_of(x)) |> 
        mutate(
          assessment = factor(assessment, assess_fct, assess_label)
        ) |> 
        pivot_wider(
          names_from = "assessment", 
          values_from = x
        ) |> 
        select(-id) |> 
        select(where(~sum(!is.na(.x))>0)) |> 
        corr_table(sample_size = F, stats = NA, numbered = T) |> 
        rename("{y}" := Variable) |> 
        flextable() |> 
        autofit()
      
    }
  ) 
```

```{r}
#| results: markup

wj_corrs_time[[1]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[2]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[3]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[4]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[5]]
```

{{< pagebreak >}}

```{r}
#| results: markup

wj_corrs_time[[6]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[7]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[8]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[9]]
```

<br/>

```{r}
#| results: markup

wj_corrs_time[[10]]
```
