---
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    fig.path: "figures/markdown/"
format: 
  docx:
    reference-doc: reference-doc.docx
    fig-dpi: 300
    fig-format: pdf
  gfm: 
    fig-format: jpeg
    fig-dpi: 300
bibliography: ../bib-files/references.bib
csl: ../bib-files/apa.csl
nocite: |
  @2005
abstract: |
  # Abstract
  The idea that some abilities might be enhanced by adversity is gaining traction. For example, research leveraging the hidden talents approach has uncovered a few specific abilities enhanced by exposure to particular forms of adversity in a given context. Yet, in order for a field to grow, we must not dig too deep, too fast. In this paper, we compliment adaptation-based research with principled exploration. To do so, we draw on the basic insights of adaptation-based research: 1) enhanced performance manifests within individuals and 2) reduced and enhanced performance can co-occur. Although commonly assumed, these assertions are rarely tested. To do so, a variety of ability measures are needed that examine relative performance differences. However, rather than using adaptive-logic to predict which abilities are enhanced or reduced, we develop statistical criteria to help interpret three different data patterns: reduced, enhanced, and intact performance. We use these criteria to analyze data from the Study of Early Childcare and Youth Development (SECCYD) to examine how adversity shapes within-person performance across 10 abilities in the Woodcock Johnson Cognitive and Achievement test batteries. Our goals are to document adversity-shaped cognitive profiles, identify possible drivers of reduced overall performance, map out sets of 'intact' abilities, and discover new enhanced abilities. We argue that principled exploration with clear criteria can help break new ground, re-map old territory, and fuel theory development. Our approach thus offers a valuable complement to the adaptive-logic approach that has dominated this emerging area of research to date.
---

```{r}
#| label: setup
#| include: false

# Load libraries
library(tidyverse)
library(patchwork)
library(flextable)

# Custom functions
source("../scripts/0-corr_table.R")

# Load staged results
load("r-objects.Rdata")

# ggplot2 theme
theme_set(
  theme_bw() +
    theme(
      axis.line.y       = element_line(),
      axis.text.y       = element_text(size = rel(1.1)),
      axis.title.y      = element_text(size = rel(1.25), margin = margin(1,0,0,0,"lines")),
      axis.ticks.y      = element_line(),
      axis.text.x       = element_text(size = rel(1.1)),
      axis.title.x      = element_text(size = rel(1.25), margin = margin(1,0,0,0,"lines")),
      axis.line.x       = element_line(),
      panel.border      = element_blank(), 
      panel.spacing.y   = unit(0.5, "lines"),
      plot.margin       = margin(.25,.25,.25,.25,"lines"),
      plot.background   = element_rect(color = NA),
      plot.title        = element_text(size = rel(1.25), hjust = 0.5, margin = margin(0,0,.5,0, "lines")),
      plot.subtitle     = element_blank(),
      panel.grid        = element_line(color = NA),
      strip.background  = element_blank(), 
      strip.placement   = "outside",
      strip.text        = element_text(size = rel(1), angle = 0)
    )
)

## Table settings
# set up flextable for tables
set_flextable_defaults(
  font.family = "Times", 
  font.size = 10,
  font.color = "black",
  line_spacing = 1,
  padding.bottom = 1, 
  padding.top = 1,
  padding.left = 1,
  padding.right = 1
)

```

{{< pagebreak >}}

# How does adversity relate to performance across different abilities in the same person?

Developmental science commonly asserts that adversity-exposure during
development reduces cognitive performance---a claim founded on decades
of empirical findings [@duncan2017a; @farah2006; @hackman2010;
@mclaughlin2019; @raby2015; @fraley2013a]. In recent years, however,
adaptation-based frameworks, rooted in the idea that adversity might
enhance certain abilities, have complemented this work---and it is
gaining traction [@ellis2017; @ellis2022; @frankenhuis2020;
@frankenhuis2013b; @frankenhuis2020a]. Since its inception, the goal of
adaptation-based frameworks has been to inspire a more well-rounded view
of adversity and its influence on abilities---one that incorporates both
the struggles and strengths of people from disadvantaged backgrounds
[@frankenhuis2013b]. As it develops further, the core task of
adaptation-based research is to "uncover a high-resolution map of
specific cognitive abilities that are enhanced as a result of growing up
under high-adversity conditions" [@ellis2017, p. 562]. To uncover this
map, researchers have used confirmatory study designs, which have
gleaned useful insights. Yet, to cultivate growth in an emerging
research program---where there is little known and much to learn---we
must not dig too deep, too soon. Without complementary approaches,
exclusive use of confirmatory designs can create tunnel vision and miss
new insights [@mcintosh2017; @roisman2021; @rozin2001; @scheel2021].

In this paper, we use a complementary approach to confirmatory research:
principled exploration. To guide our exploration, we build on two basic
insights from adaptation-based research: 1) enhanced performance
manifests within individuals, and 2) reduced and enhanced performance
can co-occur. The first insight implies we need designs and models that
can tease apart both within- and between-person performance differences.
The second suggests that, to map out more of the adversity-ability
landscape, we must examine multiple abilities measured within the same
person. Doing so will allow us to capture cognitive performance profiles
that comprise three conceptual data patterns: reduced, intact, and
enhanced performance. Past research has focused on reduced and enhanced
performance on tests of single abilities. However, we know little about
intact abilities, defined as cases where test performance is unrelated
to adversity exposure. Thus, our goal is to document adversity-shaped
cognitive performance profiles that include reduced, intact abilities,
and enhanced test performance patterns.

# **Essential Features and Empirical Insights from Adaptation-based Frameworks**

Adaptation-based research has two essential features. First, it assumes
development shapes the individual, and their abilities, to fit the local
environment [@frankenhuis2020]. Second, because environments differ in
the challenges they pose (resource-scarcity versus violence exposure),
development shapes abilities according to specific challenges. Thus,
one's abilities are thought to match the challenges of one's lived
experience. These features are useful guideposts for confirmatory
hypothesis generation. Using them as building blocks, it is possible to
construct an intuitive bridge between an ability and an environmental
challenge. For example, a researcher might identify a specific challenge
posed by a dimension of adversity (e.g., threats to safety in high-crime
neighborhoods) and an ability needed to meet the challenge (e.g.,
enhanced threat detection).

This approach is appealing because it forces researchers to be specific
and logically tie together challenges and abilities. It has also been
successful in discovering a handful of adversity-enhanced abilities,
especially in harsh and unpredictable environments. For example, past
work has proposed that constantly changing environments (i.e.,
unpredictable environments) might shape the ability to track and respond
to changing information. Using this logic, research build an intuitive
bridge between changing environments and two
abilities--attention-shifting and working memory updating---and some
empirical data are consistent with this logic [@fields2021; @mittal2015;
@young2018; @nweze2021]. However, there are two limitations to this
approach. First, previous studies are difficult to compare because they
use different measures and designs. Second, the logic behind
confirmatory hypotheses is easily flipped. For example, exposure to
unpredictable environments is thought to reduce inhibition, or the
ability to resist distractions. If opportunities are fleeting and
threats are unpredictable, inhibition is costly because focusing on
long-term goals might cause one to miss opportunities or fail to detect
a threat. But we can also assert the exact opposite. For example,
inhibition might be enhanced by unpredictable environments because
attending to every possible opportunity or threat will derail most
goal-directed actions. Thus, adaptive-logic can afford different or (in
some cases) opposing hypotheses. This does not diminish the
enterprise---empirical research is the ultimate arbiter---but there is a
risk of becoming too focused on a particular corner of hypothesis space,
when other regions would be just as reasonable to explore
[@ketelaar2000; @lewis2017; @andrews2002].

Adaptation-based research has also focused on testing content, or the
notion that performance should improve when the testing content matches
the lived experience of people exposed to adversity. For example,
studies have examined relational memory, attention shifting, and working
memory task performance using more ecologically relevant testing content
(e.g., social dominance, real-world, and socioemotional stimuli)
compared to neutral or abstract content. In some cases, ecologically
relevant content appeared to equalize performance for people exposed to
adversity, but this depends on the specific adversity measure and task
[@frankenhuis2020b; @young2022; @rifkin-graboi2021a]. Yet, in other
studies, conditions thought to be well-matched to the lived experience
of those exposed to adversity actually lower performance. For example,
youth from low socioeconomic backgrounds tend to score lower on math
items about social relations, money, and food---items thought to be
particularly relevant to lived experience---compared to other math items
[@duquennois2022; @muskens2019].

In light of various caveats, this body of work has generated at least
two empirical insights. First, although it is possible for adversity to
enhance performance between individuals (e.g., low versus high adversity
exposure), empirical findings suggest effects mostly occur within
individuals [@fields2021; @frankenhuis2020b; @young2022]. Second,
associations between specific types of adversity and enhanced
performance appears to be context specific---enhancements depend on
testing content, context, and ability type [@fields2021;
@frankenhuis2020b; @young2022; @nweze2021; @young2018; @mittal2015].
Yet, adaptation-based studies have looked for abilities in an isolated
and piecemeal fashion, in part, because confirmatory designs tend to
narrow a study's scope. This means we know little about how enhanced
abilities relate to broader sets of ability measures.

# **Motivating Principled Exploration**

We believe that adaptation-based frameworks can provide useful
guideposts. However, one should use shovels, not scalpels, when breaking
new ground. Emerging research programs have yet to lay basic groundwork
for testing theories, such as auxiliary assumptions or boundary
conditions [@scheel2021]. Our aim is to complement adaptation-based,
confirmatory research with principled exploration [@flournoy2020;
@rozin2001]. We see two benefits of this approach. The first is to
re-examine established patterns with a new lens. For example, both
deficit- and adaptation-based perspectives assume that adversity should
reduce performance on standard assessments of cognitive ability
[@ellis2022; @frankenhuis2020; @hackman2010; @mclaughlin2019;
@ursache2016]. Yet, these tests are often comprised of many different
subtests, and individual tests may show unique patterns that diverge
from widely used composite scores. The second is to feed theory with
useful description. One reason why we know little about broad sets of
abilities is that adaptive logic is yet to be developed for some
abilities. However, the lack of such logic this does not imply the
presence or absence of a functional link. A complementary approach is to
explore, describe, and follow up associations between adversity and
abilities to aid theory development. Therefore, we return to the map of
cognitive abilities that might be shaped by adversity and ask "what
territory needs exploration and which areas may need re-mapping?".

To carefully examine and interpret data in a principled exploration, it
is helpful to develop inferential criteria. For example, rather than
using adaptive logic to predict which abilities are enhanced or reduced,
we can ask what criteria are needed for evaluating and interpreting
different data patterns. In addition, research typically focuses on
reduced versus enhanced test performance, but performance on some tests
might remain intact (unaffected) by exposure to adversity
[@frankenhuis2020]. We know little about the intact performance of
people exposed to adversity. We also know little about the drivers of
reduced performance on broad and generic measures of ability and
achievement. For example, deficit approaches have collapsed many
abilities into composites and find that adversity exposure is associated
with reduced performance [@raby2015; @fraley2013a]. However, one
possibility is that a smaller set of specific performance measures are
driving effects. In total, there is still much to learn about how
adversity shapes cognitive abilities. Principled exploration can
complement confirmatory research in drawing this map, especially in the
early stages of a new field.

# The Current Study

We conduct a principled exploration of how adversity relates to
performance on a widely-used cognitive achievement battery using
longitudinal, prospective data from the Study of Early Childcare and
Youth Development (SECCYD). Drawing on the general insights of
adaptation-based research, we employ a within-person performance design
to explore performance across 10 abilities. This design allows us to
assess how exposure to each measure of adversity is associated with
relative performance differences across many abilities (see Figure 1).
In other words, we can compare specific abilities (e.g., short-term
memory performance) to overall performance (within-person average
performance on all tests) to get a clear picture of how enhanced and
reduced performance manifest in parallel within an individual.

We focus on adversity measures of two constructs: environmental
harshness and unpredictability. We focus on these constructs because
they feature often in adaptation-based research on cognitive abilities
[@ellis2017; @ellis2022; @fields2021; @frankenhuis2020; @mittal2015;
@young2018; @young2022]. Conceptually, harshness is defined as external
causes of mortality-morbidity and unpredictability is defined as random
variation in harshness over space and time [@ellis2009]. To measure
harhness, studies typically use socioeconomic indices, such as income
[@belsky2012; @simpson2012; @zhang2022; @li2018; @doom2016; @doom2022;
@szepsenwol2015; @szepsenwol2019; @sung2016; @hartman2018a]. To measure
unpredictability, studies have used a wide variety of approaches [see
@young2020], including counting family transitions and computing
variability in income scores [@li2018; @belsky2012; @hartman2018a].

We use both previously-used (i.e., income for harshness; family
transitions and income variability for unpredictability) and unexplored
measures for both. Unexplored measures include neighborhood disadvantage
(mean for harshness and variability for unpredictability). We leverage
data from the 1990 Census about the broader ecological context, which
has been used to measure the neighborhood context in the SECCYD
previously [@bleil2021; @bleil2021b].

We outline two sets of criteria for evaluating results. First, our
expectations change according to the conceptual framework. For example,
from a traditional deficit perspective, we should expect negative
overall effects of adversity. Performance on subtests should closely
match the overall effect. In contrast, from an adaptation-based
perspective, we expect an overall negative effect but performance on
some subtests is either less reduced, intact, or even enhanced.

Our second set of criteria are statistical. Our modeling strategy allows
us to quantify performance as a function of adversity in two ways.
First, we can test whether the effect of adversity on each subtest is
different from zero using a simple slopes test. A positive and negative
effect suggests enhanced and reduced performance, respectively. Second,
we compare subset performance (simple slope) against overall performance
(main effect of adversity across all tests), which is measured by the
interaction between subtest category and adversity. This interaction
term indicates whether performance is significantly more negative, less
negative, or even positive compared to overall performance. For both
types of effects, we can determine if they are practically equivalent to
either zero (simple effect) or overall performance (main effect).
Subtest performance is intact when the effect of adversity on a subtest
is practically equivalent to zero. Using these criteria, we position
ourselves to identify the drivers of reduced overall cognitive
performance, map out sets of 'intact' cognitive abilities, and discover
(possible) enhancements.

![Figure 1. Conceptual visualization of Woodcock Johnson statistical
models. A) is the main effect of adversity on overall performance. B) is
the main effect of a subtest, which reflects the average performance on
a subtest. C) is the simple effect (slope) of adversity for a particular
subtest. D) is the interaction effect that measures the difference
between A and C. A significant simple effect means the C ≠ 0 and a
significant interaction means A ≠ C. Put differently, when C is
significant, it means that adversity is associated with performance on a
subtest. When D is significant, it means that the association between
adversity and a subtest (C) is different than the association between
adversity and the overall effect
(A).](figures/fig1-conceptual.jpg){fig-align="center"}

\

# Method

## Participants

Families were initially recruited for the NICHD SECCYD in 1991. A total
of 1364 families met all the prescreening criteria, namely that mothers:
were age 18 or older, did not plan to move, had a newborn without any
known disabilities (and could leave the hospital within one week), had
no history of substance abuse, could speak English, and lived within one
hour driving distance from the research lab and were in a relatively
safe neighborhood (NICHD ECCRN, 2005). More information about
recruitment and selection procedures is available from the study (see
<https://www.icpsr.umich.edu/web/ICPSR/series/00233>). The current
analyses included participants with non-missing data on most predictors
and outcome variables through age 15 (*N* = `r intext$n`).

## Measures

### **Cognitive Ability Test Battery**

We used the Woodcock-Johnson (WJ) Cognitive and Achievement standardized
test battery to examine performance across 10 subtests [@woodcock1990;
@woodcock1990a]. The SECCYD administered the WJ five times: in the 54
month, 1^st^ grade, 3^rd^ grade, 5^th^ grade, and 15-year assessments.

There are two WJ test batteries, the cognitive and achievement tests.
The WJ cognitive test includes the Memory for Names, Memory for
Sentences, Verbal Analogies, Incomplete Words, and Picture Vocabulary
subtests (described later). The WJ achievement battery includes
Letter-Word Identification, Passage Comprehension, Calculations, Applied
Problems, and Word Attack subtests (described later).

For all tests, we analyzed standard scores, which are equivalent to IQ
scores (e.g., *M* = 100, *SD* = 15). Using standard scores for subtests
puts all tests on the same scale to facilitate comparison (see Figure
2). For each subtest, we averaged standard scores over time to create
one score per subtest, per participant. However, the specific set of
subtests administered at each assessment varied (see Figure 2). For
example, the Verbal Analogies test was measured at grade three and age
15 whereas Passage Comprehension was measured at grades 3, 5, and age 15
(see Table 1). Thus, to create overall scores for each subtest, we
averaged over all time points available for each subtest (see
<https://tinyurl.com/seccyd-wj-agg-dvs> for code).

```{r}
#| label: Figure2
#| fig-cap: |
#|   _**Figure 2.**_ WJ subtest standard scores across assessments. Different sets of subtests were administered at each asessment. Scores were averaged over assessments to create an overall subtest score. Vertical histograms reflect distributions of overall scores per subtest. Gray horizontal lines are sampe average scores for all subtests (e.g., overall WJ score).
#| fig-width: 4.5
#| fig-height: 7.75

fig2
```

**Picture Vocabulary.** This subtest measures verbal comprehension and
crystallized knowledge. The test contains 58 items requiring
participants to view and name familiar and unfamiliar objects. The test
was administered five times: at 54 months, grades 1, 3, 5, and at 15
years. Higher scores indicate more verbal comprehension and more
crystallized knowledge.

**Verbal Analogies.** This subtest measures the ability to reason about
analogies between relatively simple words. Although the words remain
simple, relations between words increase in complexity of over the test
items. The test contains 35 items and was assessed twice: at grades 3
and 5. Higher scores indicate more reasoning and more
verbal/crystallized knowledge.

**Passage Comprehension.** This subtest test measures the ability to
read a short passage and name an appropriate key word that is missing.
The test contains 43 items and was administered three times: at grades
3, 5, and at age 15. Higher scores indicate more vocabulary,
comprehension, and reading skill.

**Applied Problems.** This subtest contains a set of practical math
problems. Participants must read and identify a strategy for solving the
problem and execute simple arithmetic calculations. The test contains 60
items and was administered five times: at the 54-month, 1^st^, 3^rd^ and
5^th^ grade, and 15-year assessments. Higher scores indicate more
practical math and problem-solving skill.

**Calculations.** This subtest required participants to solve
traditional math problems containing addition, subtraction,
multiplication, division, and different combinations of each. The test
also includes some geometry and trigonometry problems. Some items
require logarithmic operations and calculus. The test contains 58 items
and was administered twice: at the 3^rd^ and 5^th^ grade assessments.
Higher scores indicate more mathematical/quantitative skill.

**Memory for Names.** This subtest is an auditory-visual association
test. It requires participants to learn a set of 'space creatures' and
their names. After learning a set of creature-name pairs, participants
are presented with nine creatures and must identify which were just
shown and which were shown previously. The test difficulty is controlled
by (decreasing) increasing the create-name pairs presented in each set.
The test contains 72 items and was administered twice: at the 1^st^ and
3^rd^ grade assessments. Higher scores indicate more visual-auditory
association and long-term memory skill.

**Incomplete Words.** This subtest measures the ability to listen to
words containing missing phonemes and complete the word. The test
contains 40 items and was administered twice: at the 54 month and 1^st^
grade assessments. Higher scores indicate more auditory processing
skill.

**Memory for Sentences.** This subtest measures the ability to listen to
and remember words, phrases, and sentences. The words, phrases, and
sentences are played on an audio tape and participants must recall as
many as possible. The test contains 32 items and was administered three
times: at the 54-month, 1^st^ grade, and 3^rd^ grade assessments. Higher
scores indicate more short-term memory skill.

**Letter-word Identification.** This subtest measures reading and
pronunciation ability. Participants must initially read letters and then
words, which gradually increase in difficulty. The test contains 57
items and was administered four times: at the 54-month, 1^st^, 3^rd^,
and 5^th^ grade assessments. Higher scores indicate more verbal
knowledge.

**Word Attack.** This subtest measures the ability to pronounce
unfamiliar words. Participants must read aloud phonetically logical but
nonsense or infrequent words. It contains 30 items and was administered
twice: at the 1^st^ and 3^rd^ grade assessments. Higher scores indicate
more auditory processing and linguistic structural analysis knowledge
and skill.

\

```{r}
#| tab.id: table1
#| results: markup

table1 |> 
  flextable() |> 
  autofit() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  add_header_row(
    values = " ",
    colwidths = 11
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 1. Bivariate correlations and descriptive statistics for WJ subtests.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 11
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 11
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_i("Note: "), as_i("* p "), "< .05, ", as_i("** p "), "< .01"), 
    part = "footer"
  )

```

\

### Indicators of Harshness

We measured environmental harshness in two ways. First, following
previous studies using data from the SECCYD, we used family
income-to-needs ratio scores from 1, 6, 15, 24, 36, and 54-month
assessments [@belsky2012; @hartman2018a; @li2018; @sung2016;
@zhang2022]. We calculated a simple average of all income-to-needs
scores across assessments to create an overall income-to-needs score
(see <https://tinyurl.com/seccyd-wj-agg-income> for code). We
reverse-scored income-to-needs mean scores to create a family income
disadvantage score where higher values indicate more disadvantage.

Second, we used data from the 1990 Census about the broader economic and
ecological context in a similar way to previous analyses of
neighborhood-level socioeconomic conditions in the SECCYD [@bleil2021;
@bleil2021b]. Specifically, addresses were tracked for each participant
over time. Each family address start and stop dates were recorded,
geocoded, and linked to the 1990 decennial Census blocks. These blocks
are the smallest Census-tracked geographical unit. For each Census
block, sociodemographic data were extracted from the Census databases to
measure neighborhood-level economic conditions for each participant. We
extracted five variables: 1) percent of people living under the poverty
line, 2) median household income, 3) Gini coefficients of income
inequality based on income frequency data, 4) percent of unemployed
individuals over 16 in the workforce, and 5) the percent of occupied
houses that were being rented. These neighborhood variables were
standardized and then averaged to create a neighborhood socioeconomic
disadvantage score for each home a participant lived in. Next, we
averaged these neighborhood scores over time (up until the 54-month
assessment). Thus, if a participant lived in two homes between birth and
the 54-month assessment, neighborhood-level variables would be
standardized and averaged within the first and second Census block, and
then averaged between them. These scores served as measures of
neighborhood socioeconomic disadvantage where higher scores indicate
higher rates of poverty, income-inequality, unemployment, lower
education, and more rental housing (see
<https://tinyurl.com/seccyd-wj-processing-census> for processing and
<https://tinyurl.com/seccyd-wj-agg-census> for aggregation).

### **Indicators of Unpredictability**

Environmental unpredictability is notoriously hard to define and measure
[@young2020]. Studies leveraging data from the SECCYD have used two
approaches. The first is track and count family transitions, including
changes in paternal figures living in the home, parental job
transitions, and residential changes [@belsky2012; @hartman2018a]. The
second approach is to quantify variability in repeated measures of
harshness indicators (e.g., computing variance in family income
disadvantage across time). For example, Li and colleagues [-@li2018] fit
a linear model to each participants' income-to-needs scores over time.
Then, they computed the residual variance around participant-level
linear trends in income-to-needs to create an income variability score.
In the current study, we compute unpredictability scores using both
approaches and extend the Li and colleagues [-@li2018] approach to the
neighborhood-level Census block data.

To calculate family transitions, we computed the number of paternal
figure changes (father figures moving in and out of the home), mother
and father (figure) job changes, and residential changes across 17
assessments from 1 to 54 months [@belsky2012; @hartman2018a]. After
computing scores across time, we standardized each variable and averaged
them together to compute an overall family transitions variable (see
<https://tinyurl.com/seccyd-wj-agg-transitions> for code).

We calculated variability scores for both family income and neighborhood
socioeconomic disadvantage. For, family income disadvantage scores, we
computed a standard deviation of all income-to-needs scores for each
participant from the 1, 6, 15, 24, 36, and 54-month assessments (see
<https://tinyurl.com/seccyd-wj-agg-income> for code). For neighborhood
socioeconomic disadvantage variability, we computed the standard
deviation of neighborhood socioeconomic disadvantage scores (see
Indicators of Harshness, above). If participants had only lived in one
Census block from 1 to 54 months, their neighborhood socioeconomic
disadvantage variability score was zero (see
<https://tinyurl.com/seccyd-wj-agg-census> for code).

### **Control Variables**

We used a standard set of three control variables typically used in
analyses of SECCYD data: 1) maternal education, 2) sex assigned at birth
(1 = female), and 3) the race/ethnicity of each child coded as
White/non-Hispanic = 0, otherwise = 1[\[WF1\]](#_msocom_1) .

 [\[WF1\]](#_msoanchor_1)Some readers may criticize this choice. Perhaps
we should motivate it or discuss it as a limitation later (adding a
parenthetical here stating that we will do so).

```{r}
#| tab.id: table2
#| results: markup

table2 |> 
  flextable() |> 
  autofit() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  add_header_row(
    values = " ",
    colwidths = 6
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 2. Bivariate correlations and descriptive statistics for adversity measures.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 6
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 6
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_i("Note: "), as_i("* p "), "< .05, ", as_i("** p "), "< .01"), 
    part = "footer"
  )
```

# Results

## Preregistration, Statistical Power, and Computational Reproducibility

We preregistered this study using a template for secondary data analysis
[@akker2021]. The preregistration document and its entire version
history was tracked on GitHub (see
<https://tinyurl.com/seccyd-wj-prereg> for the document and
<https://tinyurl.com/seccyd-wj-prereg-history> for revision history).

We also conducted a power analysis as part of our preregistration (see
<https://tinyurl.com/seccyd-wj-power> for write up and see
<https://tinyurl.com/seccyd-wj-power-code> for code). In short, we used
a simulation approach to conduct power analyses. Although we simulated
adversity scores, we used the actual WJ test scores from SECCYD data
used in this study. Simulations showed that, with a sample size of (*N*
= `r intext$n`), the smallest interaction effect we can detect is
$\beta$ = -.075 (or .075) with 90% power, if error is small. When error
is larger, we can detect the same effect size with only 65% power.
However, even with larger error, we can detect a $\beta$ = -.10 (or .10)
with 83% power.

All relevant files (data processing, analysis code, manuscript etc.) for
this project are tracked on GitHub (see
<https://tinyurl.com/seccyd-wj>), including data needed to reproduce all
results (see <https://tinyurl.com/seccyd-wj-data>). Raw data (data
provided by SECCYD) is only available via Inter-university Consortium
for Political and Social Research (ICPSR, see
<https://www.icpsr.umich.edu/web/pages/>). However, documentation for
the study is free to download (see
<https://www.icpsr.umich.edu/web/ICPSR/studies/21940>), which contains
lists of raw datasets and variables. For those who have access to raw
SECCYD data, we provide a table of raw datasets and variables used in
this project (see <https://tinyurl.com/seccyd-wj-data>).

We used R, Rstudio, and Quarto to process, analyze, and report results
[@quarto; @rcoreteam2023; @positteam2023]. For reading raw SECCYD data,
used the haven and readxl R packages [@readxl; @haven]. For data
processing, visualizations, and table creation, we used the tidyverse,
sjlabelled, ggdist, ggsci, and the patchwork R packages [@tidyverse;
@sjlabelled; @flextable; @patchwork; @ggdist; @ggsci]. For analyses,
including mixed models, simple slopes, and equivalence tests, we used
lme4, faux, ggeffects, marginaleffects, and the parameters R packages
[@bates2015; @ggeffects; @marginaleffects; @parameters; @faux].

## Data Analysis Strategy and Inferential Criteria

We used a mixed effects modeling approach to analyze how adversity
relates to WJ performance. For our primary analyses, we ran one model
per adversity variable. Each model contained sex assigned at birth,
race/ethnicity, and maternal education as covariates. Adversity and
covariates were standardized or recoded to center variables at zero.

To analyze and compare WJ subtest performance with overall WJ
performance, we restructured the data so that each participant was
represented by 10 rows, one for each WJ subtest score. Then, we created
a sum-coded contrast variable for WJ subtests with 10 levels (one for
each subtest). This type of contrast sets the model intercept to the
grand mean (e.g., the mean of all subtest scores). To analyze the
effects of adversity on test performance, we entered adversity as a main
effect and the interaction between adversity and the contrast-coded
subtest variable.

A model with this structure will contain a main effect for each
covariate, a main effect of adversity, and an interaction term for each
subtest (i.e., 10 interaction terms). The main effect of adversity
reflects the association between adversity and overall WJ performance
(e.g., within-person average of all subtests; see Figure 1). Interaction
terms reflect the association between adversity and subtest performance
*compared to the main effect of adversity* (see Figure 1). That is, they
reflect the difference between the effect of adversity on overall
performance and simple effects of adversity on subtest performance (see
Figure 1). Whereas simple effects test whether an association between
adversity and subtest performance is different from zero, interaction
terms measure whether a simple effect is different from the main effect.

Using this modeling strategy, we compute three types of effect sizes: 1)
the main effect of each adversity measure (tested in separate models),
2) the interaction effect between an adversity measure and subtest, and
3) the simple effect of adversity for each subtest. We do not have
specific point or range predictions for the effect size types above.
However, we decided a priori (see preregistration at
<https://tinyurl.com/seccyd-wj-prereg>) to consider standardized
regression coefficients (i.e., $\beta$'s) of .10 (or higher) and -.10
(or lower) as meaningful. For main effects, coefficients outside this
range indicate that overall performance is meaningfully positive or
negative across levels of adversity. For interactions, effect sizes
outside these bounds indicate that associations between adversity and
subtest performance are meaningfully more negative or more positive than
overall performance. For simple effects, effects outside these bounds
indicate that the effect of adversity on a specific subtest is
meaningfully different from zero.

We are also interested in null effects. Specifically, we use equivalence
testing to determine if a given effect is practically equivalent to a
Range of Practical Significance (ROPE). We chose a ROPE falling between
$\beta$ = -.10 and $\beta$ = .10 [@lakens2018; @kruschke2018]. Although
we report standardized coefficients, we converted our ROPE to the WJ
standard score scale by multiplying the standard deviation of standard
WJ scores (*SD* = 15) by .1. This means our ROPE was -1.5 to 1.5 for
unstandardized coefficients.

To guide interpretation, we apply a set of inferential criteria for
categorizing data patterns. We are interested in three data patterns: 1)
enhanced performance, 2) reduced performance, and 3) intact performance.
We infer 'enhanced performance' when main and simple effects are
positive, statistically different from zero, and outside the ROPE. We
infer 'reduced performance' when main and simple effects are negative,
statistically different from zero, and outside the ROPE. We infer intact
performance when a main or simple effect (and its confidence bounds) is
practically equivalent to zero (i.e., falls inside the ROPE).

We use the same criteria for interaction terms with one difference.
Because interaction terms test the difference between main and simple
effects, they quantify relative performance patterns. For 'enhanced
relative performance', interaction terms must be meaningfully positive
(outside the ROPE) and statistically significant. For 'reduced relative
performance', an interaction term must be meaningfully negative (outside
the ROPE) and statistically significant. Interaction terms that are
practically equivalent to zero reflect simple effects that closely
resemble the main effect on overall performance. However, inferring
'enhanced', 'reduced', or 'intact' relative performance depends on the
size and direction of the main effect. We are particularly interested in
cases where a main effect is negative and interaction terms are
positive. This may reflect 'enhanced relative performance' (e.g.,
meaningful and significant positive interactions), or 'less reduced'
performance on a particular subtest in the context of an overall reduced
pattern of performance.

## Primary Analyses

Our primary analyses examined how indicators of harshness and
unpredictability were associated with WJ overall and subtest
performance. We ran one mixed model per indicator for a total of five
primary analyses (two for harshness and three for
unpredictability).[\[WF1\]](#_msocom_1) 

 [\[WF1\]](#_msoanchor_1)Here we can add a note about whether or not we
correct for multiple testing. We have yet to decide this. I'm not sure.
If we don't because we're doing exploratory work and are not
interpreting p-values as support for a 'hypothesis', it would be good to
make this explicit for our readers.

All analyses controlled for the main effects of maternal education,
race/ethnicity, and sex assigned at birth. Across all models, there were
main effects for both maternal education and race/ethnicity. Lower
maternal education and having a non-White racial/ethnic background was
associated with lower WJ overall performance. No model contained effects
for sex assigned at birth. Below we describe the effects of our primary
analysis predictors (see Supplement Table 1). Primary analysis code can
be found on GitHub (see <https://tinyurl.com/seccyd-wj-primary>).

### Indicators of Harshness

**Family Income Disadvantage (mean)**. Our mixed model analyzed the
effect of family income disadvantage on overall compared with subtest WJ
performance. There was a main effect of family income disadvantage such
that a higher disadvantage was associated with lower overall WJ
performance. Equivalence tests show that this overall main effect was
meaningfully negative (outside the ROPE, see Figure 3).

Interaction effects between family income disadvantage and subtests
revealed a more nuanced landscape of associations. The association
between disadvantage and performance on Passage Completion,
Calculations, Verbal Analogies, Letter-Word, Short-Term Memory, and
Unfamiliar Words subtests did not differ from the overall main effect
(see Figure 3). However, the association between disadvantage and
performance on the Picture Vocabulary subtest was significantly and
meaningfully more negative than the overall main effect (see Figure 3).
Interestingly, the association between disadvantage and performance on
the Auditory Processing, Unfamiliar Words, and Auditory-Visual
Associations subtests were significantly more positive than the overall
main effect (see Figure 3). However, equivalence tests suggest that the
disadvantage and Unfamiliar Words performance association was inside the
ROPE, and thus practically equivalent to the main effect. The
association between disadvantage and Auditory Processing and
Auditory-Visual performance were outside the ROPE.

Our simple effects analysis tested whether the associations between
family income disadvantage and subtest performance was statistically
different from zero and whether they were practically equivalent to the
ROPE (see Figure 3). Analyses revealed that the association between
family income disadvantage and each of the subtests where significantly
and meaningfully negative, except for the Auditory Processing,
Unfamiliar Words, and Auditory-Visual Associations subtests (see Figure
3). For these tests, the association between income disadvantage and
test performance was not statistically different from zero and
practically equivalent to the ROPE (see Figure 3).

Based on our inferential criteria, the main effect of family income
disadvantage suggests that higher income disadvantage was associated
with reduced overall performance. Simple effects also revealed mostly
reduced performance on each subtest. However, for the Picture Vocabulary
subtest, the income disadvantage-performance association was
significantly and meaningfully more negative than the overall pattern,
suggesting performance on this test was particularly reduced for income
disadvantaged families. Interestingly, three subtests showed relative
enhancement to the overall pattern of income disadvantage: Auditory
Processing, Unfamiliar Words, and Auditory-Visual Associations subtests.
Yet, only the associations between income disadvantage and the Auditory
Processing and Auditory Visual Associations subtest performance were
outside the ROPE. However, simple effects were not consistent with
enhancement. Instead, simple effects revealed that the income
disadvantage-performance associations between the Auditory Processing,
Unfamiliar Words, and Auditory-Visual Associations were inside the ROPE,
suggesting higher income disadvantage was associated with intact
performance on these tests.

**Neighborhood Socioeconomic Disadvantage (Mean)**. Analyses revealed a
main effect of neighborhood socioeconomic disadvantage such that a
living in a high neighborhood socioeconomic disadvantage was associated
with reduced overall WJ performance (see Figure 3). Equivalence tests
show that this overall main effect was outside the ROPE.

Interaction effects between neighborhood socioeconomic disadvantage and
subtest were varied. Associations between socioeconomic disadvantage and
subtest performance on Passage Completion, Calculations, Letter-Word,
and Short-Term Memory did not statistically differ from the overall main
effect (see Figure 3). However, neighborhood socioeconomic disadvantage
and subtest performance associations for the Picture Vocabulary, Verbal
Analogies, and Applied Problems subtests were significantly more
negative than the main effect (see Figure 3). However, equivalence tests
showed that only the association between socioeconomic disadvantage and
Verbal Analogies subtest performance was meaningfully more negative than
the main effect. Similar to the family income disadvantage analysis,
neighborhood socioeconomic disadvantage was associated with
significantly more positive performance for the Auditory Processing and
Auditory-Visual Associations compared to the overall main effect.
Equivalence tests revealed that both associations were also meaningfully
more positive, suggesting that performance on these tests were
relatively enhanced (compared to the main effect) for participants
living in socioeconomically disadvantaged neighborhoods (see Figure 3).

Simple effects revealed that higher neighborhood socioeconomic
disadvantage was associated with statistically and meaningfully negative
performance for all subtests except for the Auditory Processing and
Auditory-Visual Associations subtests. Again, for these two subtests,
performance among those living in socioeconomically disadvantaged
neighborhoods was not statistically or meaningfully different from zero.

According to our inferential criteria, results suggest that the main
effect of neighborhood socioeconomic disadvantage is consistent with
reduced overall pattern of performance. For the Verbal Analogies
subtest, high neighborhood socioeconomic disadvantage was associated
with particularly reduced performance compared with the main effect.
However, high neighborhood disadvantage and performance associations for
the Auditory Processing and Auditory-Visual Associations subtests were
consistent with relative enhancement. Similar to the family income
disadvantage results, simple effects were not consistent with
enhancement and instead revealed mostly reduced performance. For the
Auditory Processing and Auditory-Visual Associations subtests, however,
simple effects suggest that performance remained intact at higher levels
of neighborhood socioeconomic disadvantage.

```{r}
#| label: Figure3
#| fig-width: 6
#| fig-height: 6.25
#| fig-cap: |
#|   **Figure 3.** Results of models testing the effect of family and neighborhood economic disadvantage on WJ performance. The top and bottom rows depict family and neighborhood socioeconomic disadvantage, respectively. The left column plots the overall slope (thick black lines) against the subtest slopes across low to high socioeconomic disadvantage Unfaded and faded lines are practically inequivalent and equivalent to the overall slope, respectively. The middle and right columns show interaction and simple effects. Black horizontal lines are the main effect and zero for interactions and simple effects, respectively. The gray ribbon reflects the ROPE. Solid points indicate interactions and simple effects that are practically equivalent to the ROPE. Hollow points reflect interaction and simple effects that are outside the ROPE. Statistical significance for interactions (tested against the main effect) and simple effects (tested against zero) are flagged with significance stars. \ 
#|   \*\*\* *p* < .001,  \*\* *p* < .01,  \* *p* < .05


theme_set(
  theme_light() +
    theme(
      text = element_text(size = 11),
      title = element_text(size = 10, hjust = .5),
      axis.line = element_line(),
      panel.border = element_blank(),
      panel.background = element_rect(color = NA),
      panel.grid = element_blank(),
      plot.background = element_rect(color = NA),
      plot.title = element_text(hjust = .5, face = "bold"),
      strip.background = element_rect(color = NA, fill = NA),
      strip.text = element_text(color = "black", hjust = 0.5, face = "bold.italic")
    )
)

fig3
```

### Indicators of Unpredictability

**Family Transitions**. Our analysis of family transitions revealed no
main effect on overall WJ performance. The main effect also fell inside
the ROPE range, suggesting that overall performance was not associated
with exposure to more family transitions (see Figure 4).

Three interaction terms were statistically significant: Calculations
(more negative), Auditory Processing (more positive), and Audio-Visual
Associations (more positive). However, only the association between
family transitions and performance on the Calculations was meaningfully
different from the main effect (see Figure 4).

Simple effects showed that exposure to family transitions was not
associated with subtest performance, except the Calculations and Applied
Problems subtests. For Calculations, exposure to more family transitions
was associated with significantly and meaningfully lower performance.
For Applied Problems, more family transitions were associated with
meaningfully lower performance, but this difference was not
statistically different from zero (i.e., the association was not
significant but was outside the ROPE).

Our inferential criteria suggest that exposure to family transitions was
associated with intact overall WJ performance. Simple effects suggest
that performance on most subtests was also largely intact among those
exposed to family transitions. However, for the Calculations subtest,
more family transitions were related to a pattern of reduced
performance.

**Family Income Variability (*SD*).** Models unpacking the effect of
family income variability on WJ overall and subtest performance yielded
surprising results. Specifically, the directions of all effects were
opposite to analyses using family income average scores. For subtests
that showed reduced performance at high *mean* levels of family income
disadvantage, we found enhanced performance at high levels of
*variability* in family income. We believe such effects are driven by
the fact that family income disadvantage mean and variability scores are
strongly negatively related (*r* = `r intext$incm_incsd`), which has
been reported before elsewhere [@li2018]. That is, families experiencing
more income disadvantage tended to experience less income variability.
Put differently, richer families were more likely to experience income
fluctuations.

This raises questions about using family income variability as an
indicator of adversity. In most empirical cases, higher levels of
harshness are associated with higher levels of unpredictability. Yet
here, income variability and average income are correlated in the
opposite direction. One possibility is that it matters how variability
scores are computed over repeated measures of income. Thus, to unpack
this issue, we conducted a set of secondary analyses that use different
methods for computing variability over income-to-needs scores. We report
analyses using different methods for quantifying variability in our
Secondary Analyses (see <https://tinyurl.com/seccyd-wj-update1> for the
update to our analysis plan).

**Neighborhood Socioeconomic Variability**. In contrast to family income
variability, more neighborhood socioeconomic variability was related to
higher average neighborhood socioeconomic disadvantage. That is,
families living in more socioeconomically disadvantaged neighborhoods
(more harsh) were more likely to experience variability in neighborhood
economic disadvantage (more unpredictable) from one to 54 months (*r* =
`r intext$neighm_neighsd`). Additionally, the associations between
average and variability scores were moderate rather than strong (see
Table 2).

There was no main effect of neighborhood socioeconomic variability on
overall WJ scores (see Figure 4). There was only one significant
interaction with subtest performance. High neighborhood socioeconomic
variability was associated with higher Audio-Visual Associations
performance compared to overall performance. However, this effect was
inside the ROPE, suggesting it was not meaningfully different from the
overall effect. In addition, simple effects showed that high
neighborhood socioeconomic variability was not associated with
performance on any subtest and all simple effects were inside the ROPE.

Based on our inferential criteria, high neighborhood socioeconomic
variability was associated with intact performance for overall and
individual subtest performance.

```{r}
#| label: Figure4
#| fig-width: 6.5
#| fig-height: 6.25
#| fig-cap: |
#|   **Figure 4.** Results of models testing the effect of family transitions and neighborhood socioeconomic variability on WJ performance. The top and bottom rows reflect family transitions and neighborhood socioeconomic variability, respectively. The left column plots the overall slope (thick black lines) against the subtest slopes across low to high unpredictability. Unfaded and faded lines are practically inequivalent and equivalent to the overall slope, respectively. The middle and right columns show interaction and simple effects. Black horizontal lines are the main effect and zero for interactions and simple effects, respectively. The gray ribbon reflects the ROPE. Solid points indicate interactions and simple effects that are practically equivalent to the ROPE. Hollow points reflect interaction and simple effects that are outside the ROPE. Statistical significance for interactions (tested against the main effect) and simple effects (tested against zero) are flagged with significance stars. \ 
#|   \*\*\* *p* < .001,  \*\* *p* < .01,  \* *p* < .05

fig4
```

## Secondary Analyses

Our primary analyses examining family income variability raised
questions about its validity as an adversity measure. Our secondary
analyses were designed to address this issue and explore different
methods of computing variability scores (see
<https://tinyurl.com/seccyd-wj-update1> for the secondary analysis
plan).

We computed three types of variability scores over the income-to-needs
data. The first was identical to our primary analyses; we computed a
within-person standard deviation of income-to-needs from 1 to 54 months.
Second, we computed residual standard deviations [@li2018]. To do so, we
fit a linear slope to each participant's income-to-needs data, extracted
residual scores, and computed the standard deviation of these residuals.

The last method computed percent change scores over each participant's
income-to-needs data. In time series analysis, percent change reflects
how much a score changes relative to the previous time point and scales
income accordingly. For example, if one's income is \$1,000 at one time
point and increases to \$1,500 at the next time point, their percent
change score would be .50 or 50% (\$500 increase is half of income at
the first time point). The percent change score is always relative to
the previous time point so if income increases another \$500 at time
point 3, the percent change score would be .33 or 33% (\$500 is 1/3 of
the second time point income of \$1,500). For low income families,
percent change scores can account for the fact that smaller income
fluctuations have a larger impact: a family with a monthly income of
\$1,500 that loses \$500 the next month (33% of their income) is
impacted more than a family earning \$5,000 a month (10% of their
income). After computing percent change scores for each assessment, we
averaged percent change scores to create a single percent change score
per participant.

Simple and residual standard deviation family income scores were
strongly related to both each other and to the average family income
disadvantage (see Table 3). However, average percent change scores were
only weakly related to income standard deviation and residual standard
deviation scores. In addition, average percent change in income scores
were weakly and positively related to mean family income disadvantage
scores (*r* = `r intext$incm_incpc`, see Table 3). That is, families
experiencing higher mean levels of income disadvantage also experienced
larger average percent changes in income over time. This aligns with
prior conceptual and empirical work that expects and finds that harsher
environments tend to be more unpredictable [@belsky2012; @brumbach2009;
@ellis2009; @simpson2012; @szepsenwol2015].

```{r}
#| tab.id: table3
#| results: markup

table3 |> 
  flextable() |> 
  border(i = 1, border.top = fp_border_default(style = "none", width = 0), part = "header") |> 
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 3. Bivariate correlations and descriptive statistics for family income variability scores.")),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) |> 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_i("Note: "), as_i("* p "), "< .05, ", as_i("** p "), "< .01"), 
    part = "footer"
  ) |> 
  autofit()
```

After computing each type of family income variability scores, we ran
three analyses with each as the primary predictor. We used the same
modeling strategy, covariates, and inferential criteria as our primary
analyses. Findings revealed similar patterns for both simple and
residual standard deviation scores: more variability in family income
was associated with enhanced performance, in contrast to the negative
associations with average family income disadvantage (see Figure 5).
Again, we believe this is an artifact of the relation between family
income average and variability scores.

In contrast, however, average family percent change in income did not
follow this pattern. Instead, higher percent changes in income were
consistent with intact overall WJ test performance. The only subtest
that differed from the overall effect was the Calculations subtest,
which showed that higher percent changes in income was associated with a
significant, but not meaningful, reduction in performance. Simple
effects showed higher percent changes in income were associated with
intact performance for all subtests except the Auditory Processing
subtest, which was meaningfully more positive but not statistically
different from zero.

```{r}
#| label: Figure5
#| fig-width: 6.5
#| fig-height: 7.25
#| fig-cap: |
#|   **Figure 5.** Results of models testing the effect of different family income variability scores on WJ performance. The top, middle, and bottom rows reflect simple standard deviation, residual standard deviation, and average percent change in family income from one to 54 months. The left column plots the overall slope (thick black lines) against the subtest slopes across low to high variation in family income Unfaded and faded lines are practically inequivalent and equivalent to the overall slope, respectively. The middle and right columns show interaction and simple effects. Black horizontal lines are the main effect and zero for interactions and simple effects, respectively. The gray ribbon reflects the ROPE. Solid points indicate interactions and simple effects that are practically equivalent to the ROPE. Hollow points reflect interaction and simple effects that are outside the ROPE. Statistical significance for interactions (tested against the main effect) and simple effects (tested against zero) are flagged with significance stars. \ 
#|   \*\*\* *p* < .001,  \*\* *p* < .01,  \* *p* < .05

fig5
```

# Discussion

{{< pagebreak >}}

# References

::: {#refs}
:::
