---
title: "Preregistration Update - Income-to-Needs Follow up"
format: gfm
bibliography: ../../bib-files/references.bib
csl: ../../bib-files/apa.csl
link-citations: true
execute: 
  echo: true
  warning: false
---

```{r}
#| label: setup
#| message: false

# Libraries
library(tidyverse)
library(broom)

# set directory to root directory
knitr::opts_knit$set(root.dir = "../../")
```

*Last updated `r format(Sys.time(), "on %A, %B %d, %Y at %I:%M %p")`*

## Background

Our primary analyses revewaled a strange result for our model of income-to-needs variablity. This update is to report a plan for following up this result with a set of secondary analyses.

Our model unpacking the effect of income-to-needs variablity on WJ overall and subtest performance raised questions about the validity of income-to-needs variability scores as indicators of unpredictability. In short, simple effects suggest that the subtests that were reduced by income-to-needs mean scores were enhanced in analyses using income-to-needs variablity. That is, more poverty was related to less income-to-needs variablity. Put differently, richer families were more likely to experience income fluctuations.

This is odd because it suggests that high poverty and high poverty variability show opposite effects. Moreover, neighborhood poverty and neighborhood poverty variability do not follow this pattern.

We belieive such effects are driven by the fact that income-to-needs mean and variability scores are strongly related (see below), which has been reported elsewhere [@li2018]. We also knew about this correlation before conducting our anlaysis. However, previous literature had documented this issue and it was not entirely clear what the consequence of this correlation would be for the current study. 

## Update: Secondary Analysis

```{r}
#| label: data

# Load and source relevant files
load("data/analysis-ivs.Rdata")
source("scripts/0-corr_table.R")

# ggplot2 theme
theme_set(
  theme_light() +
    theme(
      text = element_text(size = 14),
      title = element_text(size = 14, hjust = .5),
      axis.line = element_line(),
      panel.border = element_blank(),
      panel.background = element_rect(color = NA),
      panel.grid = element_blank(),
      plot.background = element_rect(color = NA),
      plot.title = element_text(hjust = .5, face = "bold"),
      strip.background = element_rect(color = NA, fill = NA),
      strip.text = element_text(color = "black", hjust = 0.5, face = "bold.italic")
    )
)
```

We plan to conduct a set of secondary analyses that use different methods for computing income-to-needs variability to unpack this issue. In our primary analyses, we used simple standard deviations of income-to-needs scores from 1 to 54 months. Li and colleagues [-@li2018] used residual variance scores. These scores are nearly identical and correlate the same with average income-to-needs scores (see below).

To handle the strong relation between average and variability scores, we computed average percent change scores. In time series analysis, percent change reflects how much a score changes relative to the previous point. For example, if one's income is \$1,000 at timepoint 1 and increases to \$1,500 at timepoint 2, their percent change score would be .50 or 50% (\$500 increase is half of income at timepoint 1). The percent change score is always relative to the previous timepoint so if income increases another \$500 at timepoint 3, the percent change score would be .33 or 33% (\$500 is 1/3 of timepoint 2 income of \$1500).

We computed all three types of income variability scores: simple standard deviation, residual variance, and percent change. Both standard deviation and residual variance scores correlate strongly with mean scores and are nearly identidical to each other. However, percent change scores only weakly correlate with mean scores and the sign of this correlation flips. That is, higher income is related to lower average percentage change scores.

```{r}
#| label: compute_scores

# Linear model function
incnt_linear <- function(df){
  lm(incnt ~ as.numeric(assessment), data=df)
}

# Grab only income-to-needs scores
new_incnt_data1 <- 
  ivs_data2 |> 
  filter(
    assessment %in% c("01", "06", "15", "24", "36", "54")
  ) |> 
  select(id, assessment, incnt)

# Nest data to apply linear function to each observation
new_incnt_data2 <- 
  new_incnt_data1 |> 
  drop_na(incnt) |> 
  nest(.by = id) |> 
  mutate(
    incnt_lm        = map(data, incnt_linear),
    incnt_mean      = map_dbl(data,\(x) mean(x$incnt, na.rm = T)),
    incnt_sd        = map_dbl(data,\(x) sd(x$incnt, na.rm = T)),
    incnt_coefs     = map(incnt_lm, tidy),
    incnt_stats     = map(incnt_lm, glance),
    incnt_intercept = map_dbl(incnt_coefs, \(x) unlist(x[1,"estimate"])),
    incnt_slope     = map_dbl(incnt_coefs, \(x) unlist(x[2,"estimate"])),
    incnt_deviance  = incnt_stats |> map_dbl("deviance"),
    incnt_sigma     = incnt_stats |> map_dbl("sigma"),
    incnt_pc        = map_dbl(data,\(x) mean(abs(x$incnt - lag(x$incnt))/lag(x$incnt), na.rm = T)),
    incnt_n         = map_dbl(data, nrow),
    .by = id
  )

new_incnt_data2 |> 
  select(matches("(mean|sd|sigma|pc|_n)$")) |> 
  corr_table(
    numbered = T,
    c.names = c("Income Mean", "Income SD", "Income Residual Variance", "Income Average Percent Change", "N scores"),
    stats = c("n", "mean", "sd", "min", "median", "max", "skew","kurtosis")
  ) |> 
  knitr::kable()
```

And this can be visualized by drawing scatter plots of each variability scores against average scores:

```{r}
#| label: scatterplots
#| fig-align: center
#| fig-dpi: 300
#| fig-width: 4
#| fig-height: 6

new_incnt_data2 |> 
  select(id, matches("(mean|sd|sigma|pc)$")) |> 
  pivot_longer(c(-id, -incnt_mean), names_to = "var_type", values_to = "score") |> 
  mutate(var_type = factor(var_type,c("incnt_sd", "incnt_sigma","incnt_pc"),c("SD", "Residual","Percent Change"))) |> 
  ggplot(aes(x = incnt_mean, y = score)) +
  geom_point(color = "gray80") +
  stat_smooth(method = "lm", se = F, color = ggsci::pal_cosmic("hallmarks_light")(10)[4]) +
  scale_x_continuous("Income-to-Needs Mean Score") +
  scale_y_continuous("Income-to-Needs Variability Score") +
  facet_wrap(~var_type, ncol = 1, scales = "free_y")
```

## Conclusion

Based on these descriptive results, we plan to conduct our follow up analysis with these three variability scores. We will follow the same exact modeling strategy as our primary analyses.

## References