---
title: "Data"
output: github_document
link-citations: true
---

*Last updated `r format(Sys.time(), "on %A, %B %d, %Y at %I:%M %p")`*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
knitr::opts_knit$set(root.dir = "../")

library(tidyverse)
library(glue)
library(knitr)
library(htmltools)
```

This directory stores `Rdata` files containing processed data.

## Scripts Producing Analysis Data Sets

The table below contain references to the scripts that produced each:

```{r}
data_creating_scripts <- 
  list.files("scripts", pattern = "^[1-3]-(.*).R$", full.names = T, recursive = T) |> 
  map(function(x){
    processed_data <- 
      read_lines(x) |> 
      str_subset("file = .*Rdata") |> 
      str_remove_all("(.*)data/") |> 
      str_remove_all('\\"') |> 
      str_remove_all("\\)$")
    
    tibble(
      Script = x |> str_remove("(.*)/"),
      `Data Output`   = processed_data
    ) |> 
      mutate(
        Script = glue("[{Script}](https://github.com/ethan-young/seccyd-wj-subtests/tree/master/scripts/{Script})")
      )
  }) |> 
  list_rbind()

data_creating_scripts |> kable()
```

## Scripts Using Processed Data

The table below shows the scripts that use processed datasets and, for each, which dataset is loaded. 

```{r}
data_using_scripts <- 
  list.files("scripts", pattern = "^(1|2|3).*.R$", full.names = T, recursive = T) |> 
  map(function(x){
    data_reading <- 
      read_lines(x) |> 
      str_subset("list.files\\(\"data\".*\\)|load\\(\".*.Rdata\"\\)") |> 
      str_trim() |> 
      str_remove(",$") |> 
      str_extract("(data(.*).Rdata)|list.files\\(.*\\)") |> 
      str_remove("data/") |> 
      paste(collapse = "<br>")
    
    if(nchar(data_reading) > 1){
      if(str_detect(data_reading, "list.files")){
        data_reading <- 
          data_reading |> 
          rlang::parse_expr() |> 
          rlang::eval_tidy() |> 
          str_remove("data/") |>
          paste(collapse = "<br>")
      }
      tibble(
        Script = x |> str_remove("(.*)/"),
        `Data Input`   = data_reading 
      ) |> 
      mutate(
        Script = glue("[{Script}](https://github.com/ethan-young/seccyd-wj-subtests/tree/master/scripts/{Script})")
      )
    } else{
      data.frame()
    }
  }) |> 
  list_rbind()

data_using_scripts |> kable()
```

## List of Raw SECCYDS Datasets

To faciliate full computational reproducibilty for those who have access to raw SECCYD data, we created a table of raw dataset names from the SECCYD repository. Anyone wishing to reproduce data/results/figures/tables must save these raw datasets in a directory outside their project root directory. 

```{r}
data_regex <- 
  list(
    "DVs" = "Woodcock.*sav$",
    "Census IVs" = "(cens1990|res).*.sas7bdat",
    "Family IVs" = "demo(.*).sas7bdat$|fam(.*).sas7bdat$|phon(.*).sas7bdat$|f\\d\\d(a|A).*.sas7bdat"
  )

map2(names(data_regex), data_regex, function(x, y){
  the_files <- 
    list.files(path = "..", recursive = T, full.names = T) |> 
    str_subset(y) |> 
    str_remove("(.*)/") |> 
    str_remove("\\.(.*)$")
  
  tibble(
    `Variable Type` = x,
    `Raw SECCYD Data Name` = the_files
  )
  
}) |> 
  list_rbind() |> 
  kable()
```

